---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epiparameter

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/epiverse-trace/epiparameter/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/epiparameter/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/epiverse-trace/epiparameter/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/epiparameter?branch=main)
<!-- badges: end -->

`R` package with epidemiological parameter distributions for infectious diseases. Eventually it will include a range of pathogens and natural history processes, as well as functions to extract parameters from reported statistics.

## Installation

The easiest way to install the development version of `epiparameter` is to use the `devtools` package:

``` r
# install.packages("devtools")
library(devtools)
install_github("epiverse-trace/epiparameter")
library(epiparameter)
```

## Quick start

To view the available distributions, then define the desired function, e.g. to extract and plot the daily probability mass function for the incubation period of influenza H7N9:

```{r example}
# View available distributions
epiparameter::list_distributions(delay_dist = "incubation")

# Extract incubation period distribution
incubation_H7N9 <- epiparameter::epidist(
  pathogen = "influenza_H7N9", 
  delay_dist = "incubation"
)

# Plot probability mass function
plot(0:10,incubation_H7N9$pmf(0:10),xlab="time since infection")
```
