---
title: "Epidemiological parameter data sources"
output: 
  bookdown::html_vignette2:
    code_folding: show
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Epidemiological parameter data sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, load-epiparameter, echo=FALSE}
library(epiparameter)
```

`{epiparameter}` provides classes and helper functions for working with
epidemiological parameters. It is not itself the canonical store of those
parameters. This vignette describes where parameters come from, how to read
each source into an `<epiparameter>` object, and how to reach the data from
languages other than R.

## Where parameters come from

Three sources feed into `<epiparameter>` objects, and they play different
roles.

**grEPI** is the WHO Collaboratory
[Global Repository of Epidemiological Parameters](https://who-collaboratory.github.io/collaboratory-grepi-web/).
It is the long-term home for curated epidemiological parameters, is
maintained collaboratively with the WHO, and is served over a public API.

**{epireview}** is the [R package](https://mrc-ide.github.io/epireview/) from
the [Pathogen Epidemiology Review Group (PERG)](https://www.imperial.ac.uk/medicine/departments/school-public-health/infectious-disease-epidemiology/mrc-global-infectious-disease-analysis/related-initiatives/perg/),
containing parameters extracted from the literature through systematic
reviews.

**{epiparameterDB}** is the parameter library bundled with `{epiparameter}`,
installed as a dependency and loaded by default. It exists so that the package
has a small, stable, offline set of parameters to develop against: it is where
new representations and bleeding-edge features are trialled, and where
examples, tests and vignettes draw their data from. It also holds parameters
needed by {epiparameter}, other Epiverse-TRACE packages or tutorials that are
not available from grEPI or {epireview}. It is deliberately not intended to
grow into a comprehensive repository -- that role belongs to grEPI and
{epireview}.

For the contents of the bundled library, see the
[Current database](database.html) vignette, which lists every entry with its
citation.

## Reading from grEPI

`epiparameter_db()` queries the grEPI API directly when `db = "grEPI"`.

```{r, grepi, eval = FALSE}
# query grEPI over the network
ep <- epiparameter_db(
  db = "grEPI",
  disease = "Ebola disease",
  epi_name = "incubation period"
)
```

The `disease` and `epi_name` arguments filter the request, and each returned
record is converted into an `<epiparameter>` object. These chunks are not
evaluated when the vignette is built, as they require network access.

## Reading from {epireview}

Parameter tables from {epireview} are converted with `as_epiparameter()`.
The [Using {epireview} with {epiparameter}](data_from_epireview.html) article
works through this in detail.

```{r, epireview, eval = FALSE}
# a parameter table from {epireview}
params <- epireview::load_epidata("marburg")$params
ep <- as_epiparameter(params[1, ])
```

## Reading the bundled library

With no `db` argument, `epiparameter_db()` reads from {epiparameterDB}.

```{r, bundled}
ep <- epiparameter_db(
  disease = "Ebola Virus Disease",
  epi_name = "incubation period",
  single_epiparameter = TRUE
)
ep
```

Entries in {epiparameterDB} exactly reflect the literature. Information not
stated in the source paper is not imputed, either from prior knowledge -- the
vector of a disease may be well known but unstated -- or by calculating from
other reported values. This is why a field is sometimes empty for an entry
even though the information is generally known: recording only what the paper
reports keeps the provenance of every value unambiguous. Values that follow
deterministically from what was reported, such as the shape and scale of a
gamma distribution given a mean and standard deviation, are reconstructed when
the entry is read into R rather than being stored.

This convention applies to {epiparameterDB}. grEPI and {epireview} set their
own criteria for what an entry may contain, so a field left empty here is not
necessarily empty there.

## Contributing parameters

Before contributing, it is worth deciding which source an entry belongs in.
The bundled library is a development set, so parameters intended for long-term
curation are better contributed to grEPI or {epireview} directly. Entries that
support package development -- exercising a distribution family, a metadata
field, or a feature under trial -- belong in {epiparameterDB}.

Entries are added by a pull request to {epiparameterDB}, editing
`inst/extdata/parameters.json`. Each entry is a single JSON object, for
example the incubation period of influenza A (H1N1):

```json
{
  "disease": "Influenza",
  "pathogen": "Influenza-A-H1N1",
  "epi_name": "incubation period",
  "probability_distribution": {
    "prob_distribution": "weibull",
    "parameters": {
      "shape": 1.74,
      "scale": 1.83
    },
    "offset": 0
  },
  "summary_statistics": {
    "quantile_values": [3.18],
    "quantile_names": ["95"],
    "median": 1.43,
    "median_ci_limits": [1.21, 1.65],
    "median_ci": 95
  },
  "citation": {
    "author": [
      { "given": "Hiroshi", "family": "Nishiura" },
      { "given": "Hisashi", "family": "Inaba" }
    ],
    "title": "Estimation of the incubation period of influenza A (H1N1-2009) among imported cases: addressing censoring using outbreak data at the origin of importation",
    "journal": "Journal of Theoretical Biology",
    "year": 2011,
    "pmid": 21168422,
    "doi": "10.1016/j.jtbi.2010.12.017"
  },
  "metadata": {
    "units": "days",
    "sample_size": 72,
    "region": "Japan",
    "transmission_mode": "natural_unknown",
    "extrinsic": false,
    "inference_method": "mle"
  },
  "method_assessment": {
    "truncation": null,
    "discretised": false,
    "censored": true,
    "right_truncated": false,
    "phase_bias_adjusted": true
  },
  "notes": "Gamma and weibull distributions had equally good fit to the data. This entry is the weibull distribution. Weibull, exponential"
}
```

Values should be recorded as reported in the source, following the convention
described above: nothing imputed, nothing derived by hand.

`data_dictionary.json` is a JSON Schema describing every field, its type and
its accepted values, and is the authoritative reference for what an entry may
contain. Every entry is validated against it automatically by a GitHub Actions
workflow when a pull request is opened, so a malformed entry fails before
review. The
[Epiverse-TRACE contributing guide](https://github.com/epiverse-trace/.github/blob/main/CONTRIBUTING.md)
covers the mechanics of opening a pull request.

## Using the data from other languages

The bundled library is stored as JSON, so it can be read without R. Two files
are shipped in {epiparameterDB}:

- `parameters.json` -- the parameter library
- `data_dictionary.json` -- a description of each field

From R, `system.file()` gives their location on disk:

```{r, json-path, eval = FALSE}
system.file("extdata", "parameters.json", package = "epiparameterDB")
system.file("extdata", "data_dictionary.json", package = "epiparameterDB")
```

Without R, both files can be read straight from the {epiparameterDB}
repository. In Python:

```python
import json
import urllib.request

url = ("https://raw.githubusercontent.com/epiverse-trace/"
       "epiparameterDB/main/inst/extdata/parameters.json")

with urllib.request.urlopen(url) as f:
    db = json.load(f)

ebola = [
    e for e in db
    if e["disease"] == "Ebola Virus Disease"
    and e["epi_name"] == "incubation period"
]

for e in ebola:
    print(e["probability_distribution"]["prob_distribution"],
          e["summary_statistics"])
#> lnorm {'mean': 12.7, 'sd': 4.31}
#> gamma {'mean': 10.3, ...}
```

And in Julia:

```julia
using JSON, Downloads

url = "https://raw.githubusercontent.com/epiverse-trace/" *
      "epiparameterDB/main/inst/extdata/parameters.json"

db = JSON.parsefile(Downloads.download(url))

ebola = filter(
    e -> e["disease"] == "Ebola Virus Disease" &&
         e["epi_name"] == "incubation period",
    db
)

for e in ebola
    println(e["probability_distribution"]["prob_distribution"])
end
#> lnorm
#> gamma
```

Each entry is an object with `disease`, `pathogen`, `epi_name`,
`probability_distribution`, `summary_statistics`, `citation`, `metadata`,
`method_assessment` and `notes` fields. `data_dictionary.json` documents what
each of these contains.

grEPI can likewise be queried from any language, as it returns JSON over HTTP.
Its [API documentation](https://who-collaboratory.github.io/collaboratory-grepi-web/#api)
describes the available endpoints and query parameters.

## Project history

`{epiparameter}` did not always work this way.

The package began as a library of epidemiological parameters together with the
functions to use them. The intention was that it would act as a 'living
systematic review' [@elliottLivingSystematicReviews2014]: parameters would be
identified through structured literature searches, extracted, quality
assessed, and added to a library that grew over time. That process followed
established systematic review guidance, drawing on the Cochrane Handbook
[@higginsCochraneHandbookSystematic2019] and PRISMA
[@pagePRISMA2020Statement2021], and was documented in a data collation and
synthesis protocol that this vignette replaces. Community contributions were
collected through open spreadsheets and reviewed by the package maintainers;
that route has since been retired in favour of pull requests, which carry the
same automated validation as any other entry.

Two things changed that plan. First, in v0.4.0 the library was separated from
the code and moved into its own package, {epiparameterDB}. This let the data be
versioned and released independently of the functions that read it, and it let
each be licensed appropriately. An R package cannot be dual licensed, so while
the code and data lived together the parameters could not be released under
CC0. {epiparameterDB} is now licensed CC0 and {epiparameter} solely MIT.
Second, and more importantly, the WHO Collaboratory began developing grEPI as
a comprehensive and sustainably maintained repository of epidemiological
parameters, developed in collaboration with Epiverse-TRACE. Alongside it,
{epireview} was established for parameters extracted through systematic
reviews by PERG.

Maintaining a third parallel library would duplicate that effort. So
`{epiparameter}` has narrowed to what it does well: providing the classes,
methods and conversions for working with epidemiological parameters, and
interoperating with the databases that curate them. The bundled library
remains, but as a development and experimentation set rather than as a
repository intended to be comprehensive.

## References
