---
title: "Parameter extraction and conversion in {epiparameter}"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter extraction and conversion in {epiparameter}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The reporting of the distribution parameters in the epidemiological literature 
is varied. In some instances the parameters themselves will be given, in others
summary statistics of the distribution are provides, such as mean, standard 
deviation, variance, among others. 

When working with epidemiological parameters it can be useful to have the values
of the parameters for common parameterisations of the distribution. For example
shape and scale parameters for the gamma and weibull distribution and the meanlog
and sdlog parameters for the lognormal distribution. 

In {epiparameter} we provide a range of conversion and extraction functions in
order to obtain distribution parameters from summary statistics and vice versa.

Here we differentiate between conversion and extraction as what can be analystically
converted, as conversions, and those that require optimisation of parameters as
extraction. This vignette aims to describe the functionality of these conversions
and extractions provided by the {epiparameter} package.

```{r setup}
library(epiparameter)
```

## Conversions

There are two styles of conversion functions in {epiparameter}. The general conversions
which have the function name `convert_dist_*()` (where `*` and `dist` are placeholders) convert
either one set of parameters to many summary statistics (e.g. `convert_lnorm_params()`)
or possibly many summary statistics input to one set of parameters (e.g. `convert_lnorm_summary_stats()`).
The other style of function is the one-to-one conversion which has the function named `dist_param2summary_stat()`
or `dist_summarystat2param()` (where `dist`, `param` and `summary_stat` are placeholders). These functions take
two arguments as input and convert to two variables as output.

### General conversion functions


